---
title: "RPE Project Group 5"
subtitle: "impact of professionalism on perceived value of participation
format:
    html:
        theme: 
            light: flatly
            dark: Darkly
        toc: true
        toc-depth: 3
        toc-expand: 1
        toc-location: right-body
        number-sections: true
        embed-resources: true
excecute: true
    cache: true
    freeze: auto
    echo: fenced    
---

# introduction

## Problem definition

## Research Objectives

To investigate the impact of professionalism on the perceived value of participation in a education

## Importance of the Topic

## Methods

## Sampling/Data collection

## Data wrangling and preparation

### setup
```{r}
library(tidyverse)
# install.packages("haven")
library(haven)
# install.packages("sjPlot")
library(sjPlot)
# install.packages("janitor")
library(janitor)

# read data 
data <- read_sav("data/(RPE- Combined-2019-12-19.sav)") 
```

```{r}
data |>
    view_df()

data <- data |> clean_names()

data |>
    view_df()

data |> 
    count()
```
## RQ1 / H1

###  Matthew Mayer

-   **H1**: Researcher professionalism has a positive effect on participants’ perceived value of participation. **independent Variable (IV)**: Researcher Professionalism **Dependent Variable (DV)**: Perceived Value of Participation

#### IV's and DV's
- **independent Variable (IV)**: Researcher Professionalism 
- **Dependent Variable (DV)**: Perceived Value of Participation

## Plan for Data wrangling

## Plan for Data visualization

## RQ2 / H2

### Harrison Gan

-   **Research Objective**: To Find Researcher’s Professionalism

-   **Hypothesis**: Researcher’s professionalism will make people more willing to give their quality input on the survey. Research Question: How does a Researcher’s professionalism help the honesty of the research?
## IV's and DV's

## Plan for Data wrangling

## Plan for Data visualization

## RQ3 / H3

### Amir Alnajjar

-  **H1**: Higher levels of researcher professionalism are positively associated with the perceived value of participation among study participants. - Amir Alnajjar 
- **H1 (expanded)**: Participants who perceive researchers as professional (e.g., respectful, organized, and ethical) will report higher satisfaction and greater perceived value from their participation. - Amir Alnajjar

## IV's and DV's

## Plan for Data wrangling

## Plan for Data visualization

## RQ4/ H4

### Allen Cosme

-   **Research Question**: How does researchers’ professionalism influence their attitudes and behavioral intentions in conducting ethical and high-quality research?

-   **Hypotheses**: H1: Higher levels of professionalism among researchers are significantly associated with more positive attitudes toward ethical research practices. H2: Higher levels of professionalism among researchers are significantly associated with stronger intentions to follow ethical standards and maintain research quality. H3: Researchers’ positive attitudes mediate the relationship between professionalism and their intentions to conduct responsible research.

## IV's and DV's

## Plan for Data wrangling

## Plan for Data visualization

## RQ5 / H5

### Ari Nurhadi

-   **Research Objective**: To examine how researcher professionalism influences participants’ trust and willingness to provide honest and high-quality responses in a study.
-   **Research Question**: How does researcher professionalism affect participants’ trust and their likelihood of providing accurate and meaningful responses? Hypothesis:
    -   H1: Higher levels of researcher professionalism are positively associated with greater participant trust, which in turn increases the likelihood of honest and high-quality responses.

## IV's and DV's

## Plan for Data visualization

## Conclusion



