---
title: "Data Wrangling with R"
Subtitle: IBM 3202
author: "Allen Cosme"
date: 2026-03-11
format: 
    html:
        theme: cosmo
        toc: true
        toc-depth: 3
        toc-location: right
        embed-resources: true
execute: 
  freeze: auto
---

> **note** this code book will follow the structure and content for the project: https://jmjung.quarto.pub/m05-1-principles-data-wrangling-with-tidyverse-in-r/

# Overview

## Learning outcomes

by the end of this module, you will be able to:

- Describe the concept of Data Wrangling.
- Describe how Tibbles are different from data frames
- Explain how to convert wide or long data to “Tidy” data
- Explain how to merge relational data sets using join functions. (Next module)
- Explain how to use grouped mutates and filter together.
- Be familiar with major dplyr functions for transforming data.
- Create a new variable with mutate() and case_when().
- Use the pipe operator to shape the data to prepare for analysis and visualization

# Intro to data wrangling with tidyverse in R.

## loading packages

```{r}
library(tidyverse)
```
# what is tidy verse
The tidyverse is a collection of R packages that share a common design philosophy and are designed to work together seamlessly. The tidyverse includes packages for data manipulation, visualization, and modeling, among other tasks. Some of the core packages in the tidyverse include:
-  ggplot2: for data visualization
-  dplyr: for data manipulation
-  tidyr: for data tidying
-  readr: for data import
-  purrr: for functional programming
-  tibble: for data frames
-  stringr: for string manipulation
-  forcats: for working with categorical variables

## initial data prep and exploration
```{r}
#install.packages("nycflights13")
library(nycflights13)
data()
```

# data wrangling with dylyr

:::{callout-note}
For mac, use `shift + option + I` for multi cursor activation
:::

Data sets in package ‘nycflights13’:

- airlines                Airline names.
- airports                Airport metadata
- flights                 Flights data
- planes                  Plane metadata.
- weather                 Hourly weather data

```{r}
flights 

flights |>
  count(year, month, day) |>
  arrange(desc(n)) |>
  mutate(month = as_factor(month)) |>
  mutate(month = fct_reorder(month,n)) |>
  ggplot(aes(month, n, fill = year)) +
  geom_col(fill = "#FF007F", show.legend = FALSE) +
  coord_flip() +
  labs(x = "Month", 
      y = "Number of flights",
      title = "Daily flights during 2013"
      )
```

