---
title: "M05-1: Principles - Expanded In-depth Data Wrangling with Tidyverse in R"
author: "Allen Cosme"
date: "`r format(Sys.time())`"
format: 
  html: 
    toc: true
    toc-depth: 4
    number-sections: true
    code-line-numbers: true
    code-fold: false
    code-link: true
    embed-resources: true
    df-print: paged
editor: visual
execute: 
  freeze: auto
  warning: false  
  error: true
---

# Air quality

## Convert airquality to a tibble (air.tib), then to a data.frame (air.df)

```{r}
library(tidyverse)

# air qual. -> tibble
air.tib <- as_tibble(airquality)

class(air.tib)
is_tibble(air.tib) 

# tibble -> base data. frame
air.df <- as.data.frame(air.tib)

#confirm data.frame
class(air.df)
is.data.frame(air.df) # true
is_tibble(air.tib) #false
```

## Find Ozone when Temp == 97 (row + Ozone-only), using Base R and Tidyverse

```{r}
air.tib[air.tib$Temp == 97, ]
```

```{r}
air.tib %>% filter(Temp ==97)
```

```{r}
air.tib$Ozone[air.tib$Temp ==97]
```

```{r}
air.tib %>% filter(Temp == 97) %>% pull(Ozone)
```

# FLight delay

## Top 10 worst arrival delays per day → average (of those) by month

```{r}
library(tidyverse)
library(nycflights13)

# Top 10 worst arrival delays per (year, month, day),
# then average those delays by month (ignoring year/day)
monthly_avg_top10 <- flights %>%
  filter(!is.na(arr_delay)) %>%                 # remove NAs so "worst" is meaningful
  group_by(year, month, day) %>%
  slice_max(order_by = arr_delay, n = 10, with_ties = FALSE) %>%
  ungroup() %>%                                 # remove day-level grouping
  group_by(month) %>%
  summarise(
    avg_delay_top10 = mean(arr_delay),
    .groups = "drop"
  )
```

## Visualization of average delays each month

```{r}
library(ggplot2)

monthly_avg_top10 %>%
  ggplot(aes(x = factor(month), y = avg_delay_top10)) +
  geom_col() +
  labs(
    title = "Average Arrival Delay by Month (Top 10 Worst Delays Per Day)",
    subtitle = "For each day, kept only the 10 largest arrival delays; then averaged by month",
    x = "Month",
    y = "Average arrival delay (minutes)",
    caption = "Data: nycflights13::flights"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )
```

::: {.callout-note title="Key Insight" icon="true"}
This chart reflects **extreme delay behavior** rather than typical delays: it averages only the
**10 worst arrival delays per day**. Months with higher bars indicate that the *worst daily delays*
tend to be especially severe in that part of the year.

In practice, this pattern often aligns with **seasonal disruption drivers** (e.g., summer storms,
winter weather, and peak travel demand), which can amplify delay severity even if the *average flight*
is not dramatically worse.
:::

# Religion vs. income

## Explore relig_income (type, size, missing values, variables) + what we can tell

```{r}
library(tidyverse)

relig_income
class(relig_income)
dim(relig_income)
glimpse(relig_income)
names(relig_income)

# missing valise check (counts N/a per column)
colSums(is.na(relig_income))

# summary for each column
summary(relig_income)
```

## help for relig_income

```{r}
help("relig_income") # or ?relig_income
```

## Does the current data support “income bracket frequency by religion”? Is it tidy? How reshape?

:::{.panel-tabset}

## Does it support the goal?

Yes, it contains the information, but it’s not in tidy (long) form for easy plotting.

## Why not tidy right now?

- Income brackets are stored as column names, not as values in a single income variable.
- That violates tidy-data rules: one variable per column.
- Here, “income bracket” is a variable spread across many columns.

## How should it be reshaped?
Use pivot_longer() so you end up with:
- religion (category)
- income (income bracket)
- count (frequency)

That format makes grouped plots (like bar charts) straightforward.
:::

## reshape data and plotting income distribution be religion

```{r}
library(tidyverse)

relig_income_longer <- relig_income %>%
  pivot_longer(
    cols = -religion,
    names_to = "income",
    values_to = "count"
  ) %>%
  filter(income != "Don't know/refused")  
```

## redo of 3.4 as one long chain

```{r}
relig_income_longer %>%
  ggplot(aes(x = income, y = count, fill = religion)) +
  geom_col(position = "dodge") +
  labs(
    title = "Income Distribution by Religion",
    x = "Income bracket",
    y = "Count"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

# Income distribution by religion

```{r}
library(tidyverse)

relig_income %>%
  pivot_longer(
    cols = -religion,
    names_to = "income",
    values_to = "count"
  ) %>%
  filter(income != "Don't know/refused") %>%
  ggplot(aes(x = income, y = count)) +
  geom_col() +
  facet_wrap(~ religion) +
  labs(
    title = "Income Distribution by Religion (Tidied with pivot_longer)",
    x = "Income bracket",
    y = "Count"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

# Rents relationship to income in the US

##  Explore us_rent_income

```{r}
library(tidyverse)

us_rent_income
class(us_rent_income)
dim(us_rent_income)
glimpse(us_rent_income)
names(us_rent_income)

colSums(is.na(us_rent_income))

summary(us_rent_income)
```

What does the data set tell us?

- it's a tibble
- there are 5 total variables GEOID, Name, variable, Estimate, moe.
- each state is combined with a variable like income or rent
- Estimate is estimated value and moe is margin of error
- missing values may be present

## help/variable meanings

```{r}
help("us_rent_income")
```

- **GEOID** = state id (CA,WA,NV, etc)
- **name** = State name
- **variable** = which metric each row is describing (in this case, income vs. rent)
- **estimate** = estimated value of metrics for state
- **moe** = "margin of error" of estimate or uncertainty

##

**goal**: plot a relationship between income and rent across different states

**does the current table directly support our goal** 

No, since income and rent are in different rows. For a scatter plot, you want a one row per state that has both rent and income

**how can we reshape**
- use the pivot_wider function in order to make the column names:
  - estimate_income, estimate_rent
  - moe_income, moe_rent
  
doing this makes it easier to plot and analyze income against rent

## Reshape to wide + plot income vs rent

```{r}
library(tidyverse)

us_rent_income_wider <- us_rent_income %>%
  pivot_wider(
    names_from  = variable,
    values_from = c(estimate, moe),
    names_sep   = "_"
  )
```

```{r}
us_rent_income_wider %>%
  ggplot(aes(x = estimate_income, y = estimate_rent)) +
  geom_point() +
  labs(
    title = "Income vs. Rent Across States",
    subtitle = "Each point is a state (estimate values)",
    x = "Estimated income",
    y = "Estimated rent"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )
```

##  data reshape to wide + plot income v. rent

```{r}
library(tidyverse)

us_rent_income %>%
  pivot_wider(
    names_from  = variable,
    values_from = c(estimate, moe),
    names_sep   = "_"
  ) %>%
  ggplot(aes(x = estimate_income, y = estimate_rent)) +
  geom_point() +
  labs(
    title = "Income vs. Rent Across States",
    subtitle = "Pivoted wider so income and rent are columns",
    x = "Estimated income",
    y = "Estimated rent"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )
```

# one long tidyverse chain

```{r}
library(tidyverse)

nba  <- data.frame(
  player = c('A','A','B','B','C','C'),
  year   = c(1,2,1,2,1,2),
  stats  = c('22/2/3','29/3/4','18/6/7','11/1/2','12/1/1','19/2/4')
)
```

## Separate stats into points, assists, steals → nba_sep

```{r}
nba_sep <- nba %>% 
  separate(
    col =  stats,
    into = c("points", "assists", "steals"),
    sep = "/",
    convert = TRUE
  )

nba_sep
```

## Unite back into one column → mba_uni 

```{r}
nba_uni <- nba_sep %>%
  unite(
    col = stats,
    points, assists, steals,
    sep = "/"
  )

nba_uni
```

# flight departures

## non-cancelled flights -> first/last depature each day

```{r}
library(tidyverse)
library(nycflights13)
library(gt)

# 1) first/last depature for non cancelled flights
daily_dep <- flights %>%
  filter(!is.na(dep_time)) %>%
  group_by(year, month, day) %>%
  summarise(
    first_dep = min(dep_time),
    last_dep = max(dep_time),
    n = n(),
    .groups = "drop"
  )

# 2) for each month, keep the day with the latest "first departure"\

latest_first_by_month <- daily_dep %>%
  group_by(year, month) %>%
  slice_max(first_dep, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(desc(first_dep))

# 3) GT table
latest_first_by_month %>%
  gt() %>%
  tab_header(
    title = "Latest First Departure Day (per Month)",
    subtitle = "Non-cancelled NYC flights; first_dep/last_dep are dep_time (HHMM)"
  ) %>%
  cols_label(
    year = "Year",
    month = "Month",
    day = "Day",
    first_dep = "First dep (HHMM)",
    last_dep = "Last dep (HHMM)",
    n = "Flights (n)"
  ) %>%
  fmt_number(columns = n, decimals = 0)
```

:::{.callout-note}
dep_time is in HHMM “clock” format (e.g., 5 = 00:05, 2359 = 23:59). That’s why the table labels it HHMM.
:::

## flights in GT 

```{r}
library(tidyverse)
library(nycflights13)
library(gt)

top5_dest_by_carrier <- flights %>%
  filter(!is.na(dep_time)) %>%        
  count(carrier, dest, sort = TRUE) %>%
  group_by(carrier) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(carrier, desc(n))

top5_dest_by_carrier %>%
  gt(groupname_col = "carrier") %>%
  tab_header(
    title = "Top 5 Destinations by Carrier",
    subtitle = "Ranked by number of flights (NYC departures)"
  ) %>%
  cols_label(
    dest = "Destination",
    n = "Flights"
  ) %>%
  fmt_number(columns = n, decimals = 0)
```

# Solar.R 

## how much is missing in Solar.R and at what percentage?

```{r}
library(tidyverse)
air.tib %>%
  summarise(
    n_rows = n(),
    n_missing_solar = sum(is.na(Solar.R)),
    pct_missing_solar = 100 * mean(is.na(Solar.R))
  )
```

## find Solar.R mean, and replace missing Solar.R with it

```{r}
Solar_Mean <- air.tib %>%
  summarise(solar_mean = mean(Solar.R, na.rm = TRUE)) %>%
  pull(solar_mean)

Solar_Mean
```

```{r}
air.tib <- air.tib %>%
  mutate(
    Solar.R = if_else(is.na(Solar.R), Solar_Mean, Solar.R)
  )
```

## conformation of no missing Solar.R
```{r}
air.tib %>%
  summarise(n_missing_solar_after = sum(is.na(Solar.R)))
```

# monthly temprature

## option 1

```{r}
library(tidyverse)

air.tib %>%
  group_by(Month) %>%
  summarise(n = sum(Temp, na.rm = TRUE), .groups = "drop") %>%   # total monthly Temp
  mutate(
    Month = recode(as.character(Month),
                   `5` = "May", `6` = "June", `7` = "July",
                   `8` = "August", `9` = "September"),
    Month = factor(Month),
    Month = fct_reorder(Month, n, .desc = TRUE)                  # reorder by total temp desc
  ) %>%
  ggplot(aes(x = n, y = Month)) +
  geom_col() +
  labs(
    title = "Total Monthly Temperature (Sum of Daily Temps)",
    x = "Total temperature (sum of Temp)",
    y = "Month"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )
```

## Option 2

```{r}
library(tidyverse)

air.tib %>%
  count(Month, wt = Temp, name = "n") %>%                        # weighted count = sum(Temp)
  mutate(
    Month = recode(as.character(Month),
                   `5` = "May", `6` = "June", `7` = "July",
                   `8` = "August", `9` = "September"),
    Month = factor(Month),
    Month = fct_reorder(Month, n, .desc = TRUE)
  ) %>%
  ggplot(aes(x = n, y = Month)) +
  geom_col() +
  labs(
    title = "Total Monthly Temperature (Weighted Count by Temp)",
    x = "Total temperature (sum of Temp)",
    y = "Month"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )
```

# Ozone vs temprature

```{r}
library(tidyverse)

air.tib %>%
  mutate(temp_wind = Temp / Wind) %>%
  select(Ozone, Temp, Wind, temp_wind) %>%
  pivot_longer(
    cols = c(Temp, Wind, temp_wind),
    names_to = "predictor",
    values_to = "x"
  ) %>%
  ggplot(aes(x = x, y = Ozone)) +
  geom_point(alpha = 0.7, na.rm = TRUE) +
  geom_smooth(method = "lm", se = FALSE, na.rm = TRUE) +
  facet_wrap(~ predictor, scales = "free_x") +
  labs(
    title = "Ozone vs Temperature, Wind, and Temp/Wind",
    x = NULL,
    y = "Ozone"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )
```

What can we gather from the data?

- there is a positive relationship between Ozone and Temprature
- opposite is true when we change temprature to wind

- Ozone vs wind temprature looks stronger since it is combining both affects 
  - wind temprature increases when Temp is high and/or Wind is low, which are both conditions associated with higher ozone. So temp_wind acts like a simple “high-heat + low-wind” index, which tends to align more tightly with Ozone than either variable alone.