---
title: "R markdown: reproducible report in data science"
author: "Allen Cosme"
date: "`r Sys.time()`"
output: 
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
    highlight: espresso
    theme: journal
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R markdown

R markdown is a format for writing reproducible, dynamic reports with R. Use it to embed R code and results into slide shows, PDF files, HTML documents, word files and more

# Structure

1. YAML header
2. text area
3. Code chunk

# Workflow

1. Open - open a file that uses the . RMD extension.
2. Write - Write content with the easy to use R markedown syntax
3. Embed - Embed R code that creates output to include in the report
4. Render - Replace R code with its output and transform the report into a slideshow, pdf, html or MS word file

# Text Area Formating tools

plain text

end a line with two spaces to start a new paragraph

*italics* and _italics_. 

**Bold** and __bold__. 

superscript^2^. 

~~Strikethrough~~. 

Hyperlink [CCIDM](https://www.cpp.edu/cba/customer-insights-lab/index.shtml)  

# header 1
## header 2
### header 3
#### header 4
##### haeder 6

Endash --
emdash ---
inline equation $A= \pi*r^2$
slide break:

***
> block quote

1. ordered list 1
2. ordered list 2
  + sub-item 1
  + sub-item 2
  
Table header  | second header
 :----------: | :------------:
 Table cell.  | cell 2
 cell 3.      | cell 4

# YAML Header
>
A YAML header is a set of key value pairs at the start of your file. Begin and end the header with a line of 3 dashes (---). (From the rmd cheatsheet
<https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf>)

- **date**: '`r format(Sys.time())`'
  - this allows to print the current time each time you knit the file
    - can be customized in various ways
    + date+time: `r Sys.time()`
    + Day, month, date, year (e.g., Sunday, January 15, 2026): '`r format (Sys.time(), "%A, %B, %C, %D")`'
    + M/D/Y (e.g., 01/25/2026): '`r format(Sys.time(), "%m/%d/%y")`'
    
***

- **toc**: yes
      - I turned the table of contents on. This will have hyperlinks in HTML and allow renders to navigate desired sections faster
- **toc_depth**: 4
      - I want to see 4 levels of headings
_ **toc_float** : true
      - adding this move toc from the front end of the document to the left side floating when you scroll down
      - by default, the floating toc are collapsed, not showing all headings. We can turn it off using the following code
          ```
          toc_float:
            collapsed: false
          ```
- **highlight**: espresso
- **theme**: journal
- **number_sections**: yes
    - turned on automatic section numbering
    
# Embed Codes

## code chunk

### Sample

* ctrl + alt + I

```{r penguins, message=TRUE, warning=TRUE}
#use ctrl+enter to run code
#install.packages("tidyverse")
library(tidyverse)

#install.packages("palmerpenguins")
library(palmerpenguins)
penguins

dim(penguins)
install.packages("skimr", repos = "https://cloud.r-project.org")
library(skimr)
skimr::skim(penguins)

#sex distribution
penguins %>%
  count(sex) %>%
  ggplot(aes(sex, n)) +
  geom_col() 
  
# of male penquins
  sum(penguins$sex =="male", na.rm = TRUE)

#total observations
nrow(penguins)

# percentage of males
m_per <- round((sum(penguins$sex == "male", na.rm = TRUE) / nrow(penguins)) * 100, 1)
#percentage of females 
f_per <- round((sum(penguins$sex == "female", na.rm = TRUE) / nrow(penguins)) * 100, 1)
```
## inline coding

* sample characteristics: 
there are `r nrow(penguins)` penguins in the data `r m_per` % of which are males and ` r f_per` are females
***

